Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-13 text/back_end_KS.qmd
working through results

From 1980 to 2020, the population of the overall Dallas-Fort Worth metropolitan area increased substantially from approximately 2.7 Million to 6.8 Million. @fig-dallas-maps showcases all six Geomarkets experienced population growth over the three census periods. However, the most pronounced increases occured in the West of Dallas/Fort Worth Metroplex and Collin & Rockwall Counties Geomarkets, which grew from approximately 340,000 to 1.4 Million and 170,000 to 1.2 Million, respectively. These population trends, as shown in @fig-dallas-maps, suggest population growth has been partly driven by urban sprawl from the metropolitan centers - City of Forth Worth and City of Dallas - into the surrounding suburban and rural areas of the metropolitan areas. This outward expansion may also account for the comparatively larger geographic size of the West of Dallas/Fort Worth Metroplex Geomarket from 1980 to 2000.

@fig-dallas-rq1-race presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Fort Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a more balanced racial/ethnic distribution, including Dallas County (58% White, 16% Black, 19% Hispanic; 6% Asian), Irving, Arlington, and Grand Prairie (54% White, 12% Black, 26% Hispanic; 16% Asian), and City of Fort Worth (57% White, 16% Black, 22% Hispanic; 3% Asian). In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 80%, respectively. Notably, by 2000, the City of Dallas was no longer a predominantly White Geomarket, with a racial/ethnic composition of approximately 24% Black, 30% Hispanic, and 3% Asian residents.

@@

Socioeconomic demographics of median income, percent living below poverty, and educational attainment over time are presented in @fig-dallas-rq1-ses for the Dallas-Fort Worth metropolitan area. Overall, median income both within and across Geomarkets remained relatively stable across the three census periods. Between 1980 to 2000, affluence levels across Geomarkets were consistent, ranging from a low of \$82,000 in City of Fort Worth to a high of \$135,00 in the Collin & Rockwall Counties Geomarket. However, the substantial increase in median income for the Collin & Rockwall Counties Geomarket during this time period plateaued by 2020, reshuffling relatively affluence levels among Geomarkets. By 2020, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$133,000, followed by West of Dallas/Fort Worth Metroplex (\$116,000), and Dallas County (\$90,000). In contrast, median household income for City of Dallas and City of Fort Worth remained relatively stable and continued to rank at the lower levels of affluence within the metropolitan area. \[KS: NEED TO CHECK THIS; I THINK CITY OF DALLAS AND CITY OF FORT WORTH ARE MIXED UP BETWEEN GRAPH VERSUS TABLE\]

@fig-dallas-rq1-ses also illustrates changes in the percentage of residents living in poverty within Geomarkets over time, revealing shifting dynamics across Geomarkets similar to those observed in median household income. In 1980, poverty rates were comparably low in both Collin & Rockwall Counties and Dallas County, moderate in the Irving, Arlington, and Grand Prarie and West of Dallas/Fort Worth Metroplex Geomarkets, and highest for City of Dallas and City of Fort Worth Geomarkets. Overtime, poverty rates declined in Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex, while increasing for all other Geomarkets. These divergent trends resulted in a widening gap in poverty levels across the Dallas-Fort Worth metropolitan area. A similar widening disparity is evident in educational attainment. Although all Geomarkets experienced increases in the percentage of residents with a bachelor's degree or higher between 1980 and 2020, the Collin & Rockwall Counties Geomarket experienced a substantially greater increase - from 23% in 1980 to 50% in 2020- maintaining the gap with other Geomarkets in terms of educational attainment.

The Dallas-Fort Worth metroplex experienced substantial demographic growth since Geomarket boundaries were dilineated by @RN4982 in the early 1980s, leading to an urban landscape with considerable racial/ethnic diversity across multiple Geomarkets (as opposed to concentrated within a single Geomarket like Chicago) but stark socioeconomic disparities between Geomarkets. While all six Geomarkets experienced population increases, the most significant growth occurred in the outer suburban areas, reflecting ongoing urban sprawl. Such growth is largely attributed to increases in Hispanic, Black, and Asian populations leading to only two of the six Dallas-Fort Worth Geomarkets remaining predominantly White (Collin and Rockwall Counties and West of Dallas/Fort Worth Metroplex. On the other hand, the metropolitan area has experienced widening disparities in income, poverty, and education attainment between Geomarkets, especially between the historically affluent Collin and Rockwall Counties and the less economically advantaged inner-core Geomarkets (City of Fort Worth and City of Dallas). Together, these trends underscore the growingly racial/ethnic diverse but widening economic inequality within the evolving Geomarket landscape of the Dallas-Fort Worth metroplex.

@@

Prospect pools within Geomarkets from the middle range PSAT score (1190-1260) to the high PSAT score range (1270-1520) become disproportionately more Asian, while the representation of Hispanic and Black prospects decline substantially. Such shifts also shift distribution patterns of racial/ethnic diversity across Geomarkets, making City of Dallas and City of Fort Worth critical Geomarkets for including Black and Hispanic students from the resulting prospect pools. \[KS LEFT OFF HERE}

@fig-dallas-rq2-midSAT-race presents the race/ethnicity of all prospects targeted in the middle PSAT score range across Geomarket contributions. Disparities in Geomarket contributions were evident but differed for Asian, Black, and Hispanic students. For instance, The Dallas County (20% versus 13% for all prospects) and Colin and Rockwall Counties (47% versus 37% for all prospects) Geomarkets contributed a disproportionate larger share of Asian students, whereas nearly all other Geomarkets contributed disproportionately smaller shares.

@@

Lastly, @fig-dallas-rq2-midSAT-combo presents Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity in the Dallas-Fort Worth metropolitan area. Among all White prospects (n=2,094) whose contact information was purchased in this middle PSAT score range, approximately 3% were first-generation college students whose parents did not attend college, 11% were first-generation college students whose parents attended some college but did not complete their degree, and 86% were not first-generation college students. @fig-dallas-rq2-midSAT-combo shows nearly all Geomarkets contribute a nearly proportionate share of prospects by first-generation college students relative to the overall pool. The only exception is the Irving, Arlington, & Grand Prairie Geomarket contributing a relatively larger share of White first-generation college students whose parents attended some college but did not complete their degree (21% versus 11% for all prospects).

Similar patterns of one or two Geomarkets disproportionately contributing larger shares of first-generation college students was evident across other racial/ethnic categories in @fig-dallas-rq2-midSAT-combo. For instance, City of Dallas (18%) and City of Fort Worth (24%) contributed larger proportions of first-generation college students whose parents did not attend college relative to the overall pool of all prospects (8%). On the other hand, the Irving, Arlington, & Grand Prairie (23%) and Dallas Counties (26%) Geomarkets contributed disproportionately larger proportions of Black students whose parents attended some college but did not complete their degree relative to the overall pool of all prospects (17%). For Asian prospects, City of Fort Worth and Irving, Arlington, & Grand Prairie Geomarkets contributed disproportionately larger shares of first-generation college students across both parents with no college and some college. City of Dallas (38%), City of Fort Worth (39%), Irving, Arlington, & Grand Prairie (32%), Dallas Counties (28%) all contributed larger shares of Hispanic students whose parents did not attend college relative to the overall prospect pool of these first-generation college students (23%), whereas only Dallas Counties contributed to larger shares of Hispanic students who parents attended some college but did not attain their degree (26% versus 20% for all prospects).

@@

### Decentralized Diversity and Income in a Multi-Nucleo in the Los Angeles Geomarket Landscape

The 11 Geomarkets within the Los Angeles metropolitan area are illustrated in @fig-la-maps by total population across the three census periods. The Los Angeles metropolitan area is one of the only multi-nuclei models in the country that is formed when urban growth occurs around multiple centers that serve different functions rather than expanding from a single central business district core (CITE- Harris and Ullman 1945). As depicted in @fig-la-maps, Geomarkets closely follow this overall metropolitan model. Hollywood and Wilshire, South and South Central Los Angeles, and East Los Angeles Geomarkets are centrally located in the metropolitan area followed by an outer core ring of made up by the Glendale and Pasadena, San Fernando Valley-East, West Los Angeles and West Beach, South Bay, and Long Beach Geomarkets. The San Fernando Valley-West Geomarket is located in the northwest part of the region, whereas Covina and West Covina as well as the Riverside, San Bernadino, and Ontario Geomarkets are located inland.

@@

\@fig-la-rq1-race presents the racial and ethnic composition of each Geomarket within the Los Angeles metropolitan area from 1980 to 2020. Over this period, the proportion of White residents declined across all Geomarkets, while Asian and Hispanic populations generally increased. In 1980, all Geomarkets - except for East Los Angeles, South and South Central Los Angeles, and Hollywood and Wilshire- were predominantly White, with percentages ranging from 62% White in Long Beach to 78% in the San Fernando Valleys (West and East) Geomarkets. By 2020, however, only the West Los Angeles and West Beach Geomarket remained predominantly White (62%). Remanining Geomarkets exhibited a more racially/ethnic diverse population by 2020, including San Fernando Valley-West- (44% White, 4% Black, 39% Hispanic; 11% Asian Pacific Islander), San Fernando Valley -East (43% White, 4% Black, 41% Hispanic; 8% Asian Pacific Islander), Glendale and Pasadena (35% White, 3% Black, 37% Hispanic; 21% Asian Pacific Islander), South Bay (29% White, 12% Black, 33% Hispanic; 21% Asian Pacific Islander), Long Beach (26% White, 11% Black, 46% Hispanic; 13% Asian Pacific Islander), Covina and West Covina (19% White, 3% Black, 52% Hispanic; 24% Asian Pacific Islander), and Riverside, San Bernardino, and Ontario (28% White, 7% Black, 54% Hispanic; 6% Asian Pacific Islander). However, the South and South Central Los Angeles and Hollywood and Wilshire Geomarkets experienced substantial declines in the share of Black residents from 1980 to 2020, dropping from 44% to 18% and 18% to 9%, respectively.

@fig-la-rq1-ses presents trends in socioeconomic indicators for the Los Angeles metropolitan area from 1980 to 2020. While median income increased across all of the metropolitan area over the three census periods, some shifts occurred in the relative socioeconomic levels across Geomarkets. Many Geomarkets kept their relative levels of affluence from 1980 to 2020. For instance, the Covina and West Covina, South Bay, and San Fernando Valley remained among the most affluent, whereas the San Fernando Valley East, Glendale and Pasadena, Long Beach, and Riverside, San Bernardino, and Ontario Geomarkets remained at middle ranges of affluence and the Hollywood and Wilshire, South and South Central Los Angeles, and East Los Angeles Geomarkets remained at the lowest levels of affluence across the metropolitan area during this period. The West Los Angeles and West Beach Geomarket demonstrated the most substantial increase in median household income, rising from \$112,000 in 1980 to \$132,000 in 2000, and reaching \$137,000 by 2020. This sustained growth positioned it as the most affluent Geomarket in the Los Angeles metropolitan area.

@@

The Los Angeles metropolitan area exemplifies a complex multi-nuclei urban model, characterized by decentralized population growth and increasing demographic diversity across its 11 Geomarkets. From 1980 to 2020, the region’s population expanded significantly leading to overall declines in the White population and increases in Hispanic and Asian populations across Geomarkets. West Los Angeles and West Beach is the only majority White Geomarket in 2020. Socioeconomic indicators reveal enduring disparities, especially in historically lower-income Geomarkets like South and South Central Los Angeles, East Los Angeles, and Hollywood and Wilshire. Meanwhile, Geomarkets such as West Los Angeles and West Beach became increasingly affluent and highly educated, highlighting growing socioeconomic and educational divides. Given these demographic shifts within the context of a multi-nuclei urban model, the Los Angeles metropolitan area presents a distinctive Geomarket landscape case in which racial/ethnic enclaves across a range of various socioeconomic levels are distributed across multiple Geomarkets rather than concentrated in one or two.

@@

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets in the Los Angeles metropolitan areas, we analyze six orders placed by a research university filtered for prospects across the entire state of California, in the 2019-2020 high school graduating class, and by PSAT scores. Two orders indicated PSAT score thresholds ranging from 1070-1180, another two orders filtered for a 1190-1260 PSAT range, and the remaining two orders filtered for scores ranging from 1270-1520.

The racial/ethnic composition of purchased student profiles within Geomarkets in the Los Angeles metropolitan area is presented in @fig-la-rq2-race. The more than 15,000 prospects in the resulting student lists at the low PSAT score thresholds (1070-1180) reflect very similar racial/ethnic compositions within Geomarkets than overall demographic makeup of the metropolitan area in 2020 (see @fig-la-rq1-race). However, the racial/ethnic composition of prospects become less proportional to overall metropolitan patterns by Geomarket as PSAT scores increase. For instance, prospect pools at middle range PSAT scores (1190-1260) within Geomarkets become disproportionately more Asian, while the representation of Hispanic and Black prospects decline substantially. This pattern becomes most pronounced at the highest PSAT score prospect pools. This disproportionate over representation for Asian prospects in comparison to the population of residents with Geomarkets ranges from 3 percentage points in the West Los Angeles and West Beach Geomarket (16% prospects versus 13% Asian residents) to 52 percentage points in the East Los Angeles Geomarket (79% prospects versus 27% residents). @fig-la-rq2-race illustrates the share of Black prospects within these high PSAT orders declines to less than 3% within all Geomarkets in comparison to Black residents making up a range of 3% (Covina and West Covina) to 18% (South and South Central Los Angeles) of residents within Geomarkets. Similar underrepresentation patterns are evident for Hispanic porspects are evident across all Geomarkets execept for a nearly equal representation in South and South Central Los Angeles (73% prospects versus 71% Hispanic residents).

@fig-la-rq2-midSAT-race presents the race/ethnicity of all prospects targeted in the middle PSAT score range across Geomarket contributions for the Los Angeles metropolitan area. For the more than 2,500 White prospects in this pool, we see the San Fernando Valley - West (28% versus 20% for all prospects) and the West Los Angeles and West Beach (12% versus 7% for all prospects)Geomarkets contributing disproportionately larger shares. Larger disparities across Geomarket contributions were evident for Asian, Black, and Hispanic students. The Glendale and Pasadena (17% versus 14% for all prospects), East Los Angeles (11% versus 5% for all prospects) and Covina and West Covina (17% versus 11% for all prospects) Geomarkets contributed a disproportionate larger share of Asian students, whereas nearly all other Geomarkets contributed disproportionately smaller or nearly equal shares.

Representational patterns across Geomarkets are reversed for Black and Hispanic prospects in @fig-la-rq2-midSAT-race. South and South Central Los Angeles and Riverside, San Bernadino, and Ontario Geomarkets contributed larger proportions of Black and Hispanic prospects relative to their contributions to the overall prospect pool in the Los Angeles metropolitan area. On the other hand, the San Fernando Valley- West and the Glendale and Pasadena Geomarket contributed smaller proportions of Black and Hispanic Prospects. However, the West Los Angeles and West Beach (10% versus 7% for all prospects), Hollywood Wilshire (7% versus 5% for all prospects), and the Long Beach (8% versus 5% for all prospects) Geomarkets contributed a larger share of Black prospects, whereas Covina and West Covina contributed a larger share of Hispanic students. All other Geomarkets contributed nearly equal shares of Black and Hispanic prospects.

Contributions of each Geomarket to the pool of prospects from the Los Angeles metropolitan area based on first-generation college student status are presented in @fig-la-rq2-midSAT-ses. Only 889 of the more than 8,000 total prospects (11%) in this middle PSAT score range (1190-1260) pool were first-generation college students whose parents did not attend college. Only three of the 11 Geomarkets contributed substantially larger shares of these first-generation prospects: East Los Angeles (35%), South and South Central Los Angeles (47%), and Riverside, San Bernardino, and Ontario (15%). However, disproportionate contributions become more spread across Geomarkets for the nearly 1,436 of the more than 8,000 total prospects (18%) who were first-generation college students whose parents attended but did not complete their degree. Overall, six of the 11 Geomarkets contributed larger shares of these prospects: Hollywood & Wilshire (25%), East Los Angeles (26%), South and South Central Los Angeles (28%), Long Beach (25%), Covina and West Covina (22%), and Riverside, San Bernardino, and Ontario (22%). This leads to the remaining Geomarkets - San Fernando Valley- West (80%), San Fernando Valley- East (78%), Glendale and Pasadena (77%), West Los Angeles & West Beach (87%), South Bay (81%) - contributing disproportionately larger shares of not first-generation college students relative to the overall prospect pool (71%).

>>>>

From 1980 to 2020, the population of the overall Dallas-Fort Worth metropolitan area increased substantially from approximately 2.7 Million to 6.8 Million. @fig-dallas-maps shows all six Geomarkets experienced population growth over the three census periods. However, the most pronounced increases occured in the West of Dallas/Fort Worth Metroplex and Collin & Rockwall Counties Geomarkets, which grew from approximately 340,000 to 1.4 Million and 170,000 to 1.2 Million, respectively. These population trends, as shown in @fig-dallas-maps, suggest population growth has been partly driven by urban sprawl from the metropolitan centers - City of Forth Worth and City of Dallas - into the surrounding suburban and rural areas of the metropolitan areas. This outward expansion may also account for the comparatively larger geographic size of the West of Dallas/Fort Worth Metroplex Geomarket from 1980 to 2000.

[WRITING IS QUITE GOOD!I THINK THERE IS A MODEST LEAD THAT THE NEXT COUPLE PARAGRAPHS DON'T STRESS ENOUGH , WHICH IS THAT TX19/CITY OF DALLAS IS THE MOST LEAST WHITE/ASIAN AND THE POOREST]
@fig-dallas-rq1-race presents the racial and ethnic composition of each Geomarket within the Dallas-Fort Worth metropolitan area across the three census periods. In 1980, all Geomarkets were predominantly White, with White populations ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Fort Worth Metroplex. By 2000, however, Geomarkets located in the central part of the metropolitan area exhibited a more balanced racial/ethnic distribution, including Dallas County (58% White, 16% Black, 19% Hispanic; 6% Asian), Irving, Arlington, and Grand Prairie (54% White, 12% Black, 26% Hispanic; 16% Asian), and City of Fort Worth (57% White, 16% Black, 22% Hispanic; 3% Asian). In contrast, Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex Geomarkets remained predominantly White in 2000 at 78% and 80%, respectively. Notably, by 2000, the City of Dallas was no longer a predominantly White Geomarket, with a racial/ethnic composition of approximately 24% Black, 30% Hispanic, and 3% Asian residents.

@@

[CITY OF DALLAS!]Socioeconomic demographics of median income, percent living below poverty, and educational attainment over time are presented in @fig-dallas-rq1-ses for the Dallas-Fort Worth metropolitan area. Overall, median income both within and across Geomarkets remained relatively stable across the three census periods. Between 1980 to 2000, affluence levels across Geomarkets were consistent, ranging from a low of \$82,000 in City of Fort Worth to a high of \$135,00 in the Collin & Rockwall Counties Geomarket. However, the substantial increase in median income for the Collin & Rockwall Counties Geomarket during this time period plateaued by 2020, reshuffling relatively affluence levels among Geomarkets. By 2020, Collin & Rockwall Counties remained the most affluent Geomarket with a median household income of \$133,000, followed by West of Dallas/Fort Worth Metroplex (\$116,000), and Dallas County (\$90,000). In contrast, median household income for City of Dallas and City of Fort Worth remained relatively stable and continued to rank at the lower levels of affluence within the metropolitan area. \[KS: NEED TO CHECK THIS; I THINK CITY OF DALLAS AND CITY OF FORT WORTH ARE MIXED UP BETWEEN GRAPH VERSUS TABLE\]

[DON'T KNOW IF WE NEED PARAGRAPH ON POVERTY + EDUCATIONAL ATTAINMENT]. @fig-dallas-rq1-ses also illustrates changes in the percentage of residents living in poverty within Geomarkets over time, revealing shifting dynamics across Geomarkets similar to those observed in median household income. In 1980, poverty rates were comparably low in both Collin & Rockwall Counties and Dallas County, moderate in the Irving, Arlington, and Grand Prarie and West of Dallas/Fort Worth Metroplex Geomarkets, and highest for City of Dallas and City of Fort Worth Geomarkets. Overtime, poverty rates declined in Collin & Rockwall Counties and West of Dallas/Fort Worth Metroplex, while increasing for all other Geomarkets. These divergent trends resulted in a widening gap in poverty levels across the Dallas-Fort Worth metropolitan area. A similar widening disparity is evident in educational attainment. Although all Geomarkets experienced increases in the percentage of residents with a bachelor's degree or higher between 1980 and 2020, the Collin & Rockwall Counties Geomarket experienced a substantially greater increase - from 23% in 1980 to 50% in 2020- maintaining the gap with other Geomarkets in terms of educational attainment.

[DON'T THINK YOU NEED SUMMARY PARAGRAPHS LIKE THIS WITHIN SECTIONS FOR THIS METRO]The Dallas-Fort Worth metroplex experienced substantial demographic growth since Geomarket boundaries were dilineated by @RN4982 in the early 1980s, leading to an urban landscape with considerable racial/ethnic diversity across multiple Geomarkets (as opposed to concentrated within a single Geomarket like Chicago) but stark socioeconomic disparities between Geomarkets. While all six Geomarkets experienced population increases, the most significant growth occurred in the outer suburban areas, reflecting ongoing urban sprawl. Such growth is largely attributed to increases in Hispanic, Black, and Asian populations leading to only two of the six Dallas-Fort Worth Geomarkets remaining predominantly White (Collin and Rockwall Counties and West of Dallas/Fort Worth Metroplex. On the other hand, the metropolitan area has experienced widening disparities in income, poverty, and education attainment between Geomarkets, especially between the historically affluent Collin and Rockwall Counties and the less economically advantaged inner-core Geomarkets (City of Fort Worth and City of Dallas). Together, these trends underscore the growingly racial/ethnic diverse but widening economic inequality within the evolving Geomarket landscape of the Dallas-Fort Worth metroplex.

@@

Prospect pools within Geomarkets from the middle range PSAT score (1190-1260) to the high PSAT score range (1270-1520) become disproportionately more Asian, while the representation of Hispanic and Black prospects decline substantially. 
<!--OZAN CUT THIS TEXT 5/13. THINKING NOT ESSENTIAL. AND NOT SO SURE IF IT SHOWS THAT. Such shifts also shift distribution patterns of racial/ethnic diversity across Geomarkets, making City of Dallas and City of Fort Worth critical Geomarkets for including Black and Hispanic students from the resulting prospect pools. \[KS LEFT OFF HERE} -->

@fig-dallas-rq2-midSAT-race presents the race/ethnicity of all prospects targeted in the middle PSAT score range across Geomarket contributions. The bars for "All (race known)" show that Collin & Rockwall (37%) and West of Dallas/Forth Worth (23%) contribute many more students than the other four Geomarkets. Disparities in Geomarket contributions were evident but differed for Asian, Black, and Hispanic students. For instance, The Dallas County (20% versus 13% for all prospects) and Colin and Rockwall Counties (47% versus 37% for all prospects) Geomarkets contributed a disproportionate larger share of Asian students, whereas nearly all other Geomarkets contributed disproportionately smaller shares.

@@

Lastly, @fig-dallas-rq2-midSAT-combo presents Geomarket contributions to prospect pools by both first-generation college student status and race/ethnicity, which we view as the most interesting results for the Dallas-Fort Worth metropolitan area. Among all White prospects (n=2,094) whose contact information was purchased in this middle PSAT score range, approximately 3% were first-generation college students whose parents did not attend college, 11% were first-generation college students whose parents attended some college but did not complete their degree, and 86% were not first-generation college students. @fig-dallas-rq2-midSAT-combo shows nearly all Geomarkets contribute a nearly proportionate share of prospects by first-generation college students relative to the overall pool, the only modest exception being the Irving, Arlington, & Grand Prairie Geomarket.

By contrast, for non-White groups, a larger share of prospects were first-generation students. Furthermore, first generation status differed substantially across Geomarkets, especially for Hispanic and Asian students. For Hispanic students, more than 50% of prospects from the City of Dallas, City of Forth Worth, Arlington \& Grand Prairie, and Dallas County are first-generation college students. By contrast, for the more affluent Collin & Rockwall Geomarket and the West of Dallas/Fort Worth Metroplex Geomarkets, less than 30% of prospects were first-generation. For Asian prospects, City of Fort Worth and Irving, Arlington, & Grand Prairie Geomarkets contributed disproportionately larger shares of first-generation college students across both parents with no college and some college. By contrast, the more affluent Collin & Rockwall sent more Asian prospects (285) than any other Geomarket, but about 87% of these students had parents with a BA.

<!-- I ADDED THE ABOVE PARAGRAPH AND CUT THIS PARAGRAPH TO GIVE THE LEDE MORE PUNCH 
Similar patterns of one or two Geomarkets disproportionately contributing larger shares of first-generation college students was evident across other racial/ethnic categories in @fig-dallas-rq2-midSAT-combo. For instance, City of Dallas (18%) and City of Fort Worth (24%) contributed larger proportions of first-generation college students whose parents did not attend college relative to the overall pool of all prospects (8%). On the other hand, the Irving, Arlington, & Grand Prairie (23%) and Dallas Counties (26%) Geomarkets contributed disproportionately larger proportions of Black students whose parents attended some college but did not complete their degree relative to the overall pool of all prospects (17%). For Asian prospects, City of Fort Worth and Irving, Arlington, & Grand Prairie Geomarkets contributed disproportionately larger shares of first-generation college students across both parents with no college and some college. City of Dallas (38%), City of Fort Worth (39%), Irving, Arlington, & Grand Prairie (32%), Dallas Counties (28%) all contributed larger shares of Hispanic students whose parents did not attend college relative to the overall prospect pool of these first-generation college students (23%), whereas only Dallas Counties contributed to larger shares of Hispanic students who parents attended some college but did not attain their degree (26% versus 20% for all prospects).
-->

@@

### Decentralized Diversity and Income in a Multi-Nucleo Landscape

Due to the large number of Geomarkets in urban Southern California, we distinguish Los Angeles -- 11 Geomarkets, results shown here -- from Orange County (four Geomarkets) and San Diego (three Geomarkets), which are shown in the Online Appendix. The 11 Geomarkets within the Los Angeles metropolitan area are illustrated in @fig-la-maps by total population across the three census periods. The Los Angeles metropolitan area is one of the only multi-nuclei models in the country that is formed when urban growth occurs around multiple centers that serve different functions rather than expanding from a single central business district core (CITE- Harris and Ullman 1945). As depicted in @fig-la-maps, Geomarkets closely follow this overall metropolitan model. Hollywood and Wilshire, South and South Central Los Angeles, and East Los Angeles Geomarkets are centrally located in the metropolitan area followed by an outer core ring of made up by the Glendale and Pasadena, San Fernando Valley-East, West Los Angeles and West Beach, South Bay, and Long Beach Geomarkets. The San Fernando Valley-West Geomarket is located in the northwest part of the region, whereas Covina and West Covina as well as the Riverside, San Bernadino, and Ontario Geomarkets are located inland, comprising the Inland Empire.

@@

\@fig-la-rq1-race presents the racial and ethnic composition of each Geomarket within the Los Angeles metropolitan area from 1980 to 2020. Over this period, the proportion of White residents declined across all Geomarkets, while Asian and Hispanic populations generally increased. [THE CAVEAT THAT WE DON'T HAVE ASIAN FOR 1980 IS MORE IMPORTANT FOR LA] In 1980, all Geomarkets - except for East Los Angeles, South and South Central Los Angeles, and Hollywood and Wilshire- were predominantly White, with percentages ranging from 62% White in Long Beach to 78% in the San Fernando Valleys (West and East) Geomarkets. By 2020, however, only the West Los Angeles and West Beach Geomarket remained predominantly White (62%). Remanining Geomarkets exhibited a more racially/ethnic diverse population by 2020, including San Fernando Valley-West- (44% White, 4% Black, 39% Hispanic; 11% Asian Pacific Islander), San Fernando Valley -East (43% White, 4% Black, 41% Hispanic; 8% Asian Pacific Islander), Glendale and Pasadena (35% White, 3% Black, 37% Hispanic; 21% Asian Pacific Islander), South Bay (29% White, 12% Black, 33% Hispanic; 21% Asian Pacific Islander), Long Beach (26% White, 11% Black, 46% Hispanic; 13% Asian Pacific Islander), Covina and West Covina (19% White, 3% Black, 52% Hispanic; 24% Asian Pacific Islander), and Riverside, San Bernardino, and Ontario (28% White, 7% Black, 54% Hispanic; 6% Asian Pacific Islander). However, the South and South Central Los Angeles and Hollywood and Wilshire Geomarkets experienced substantial declines in the share of Black residents from 1980 to 2020, dropping from 44% to 18% and 18% to 9%, respectively.

[I THINK THE LEADS ARE THAT SOUTH CENTRAL (ESPECIALLY), EAST LA, AND WILSHIRE ARE THE POOREST. START TEHRE AND THEN GET TO OTHER RESULTS.]
@fig-la-rq1-ses presents trends in socioeconomic indicators for the Los Angeles metropolitan area from 1980 to 2020. While median income increased across all of the metropolitan area over the three census periods, some shifts occurred in the relative socioeconomic levels across Geomarkets. Many Geomarkets kept their relative levels of affluence from 1980 to 2020. For instance, the Covina and West Covina, South Bay, and San Fernando Valley remained among the most affluent, whereas the San Fernando Valley East, Glendale and Pasadena, Long Beach, and Riverside, San Bernardino, and Ontario Geomarkets remained at middle ranges of affluence and the Hollywood and Wilshire, South and South Central Los Angeles, and East Los Angeles Geomarkets remained at the lowest levels of affluence across the metropolitan area during this period. The West Los Angeles and West Beach Geomarket demonstrated the most substantial increase in median household income, rising from \$112,000 in 1980 to \$132,000 in 2000, and reaching \$137,000 by 2020. This sustained growth positioned it as the most affluent Geomarket in the Los Angeles metropolitan area.

@@

[THINK YOU CAN CUT THIS PARAGRAPH] The Los Angeles metropolitan area exemplifies a complex multi-nuclei urban model, characterized by decentralized population growth and increasing demographic diversity across its 11 Geomarkets. From 1980 to 2020, the region’s population expanded significantly leading to overall declines in the White population and increases in Hispanic and Asian populations across Geomarkets. West Los Angeles and West Beach is the only majority White Geomarket in 2020. Socioeconomic indicators reveal enduring disparities, especially in historically lower-income Geomarkets like South and South Central Los Angeles, East Los Angeles, and Hollywood and Wilshire. Meanwhile, Geomarkets such as West Los Angeles and West Beach became increasingly affluent and highly educated, highlighting growing socioeconomic and educational divides. Given these demographic shifts within the context of a multi-nuclei urban model, the Los Angeles metropolitan area presents a distinctive Geomarket landscape case in which racial/ethnic enclaves across a range of various socioeconomic levels are distributed across multiple Geomarkets rather than concentrated in one or two.

@@

To examine how the socioeconomic and racial composition of included versus excluded prospects varies when student list purchases filter on particular Geomarkets in the Los Angeles metropolitan areas, we analyze six orders placed by a research university that filtered for prospects across the entire state of California, in the 2019-2020 [NOT JUST 2019-20. REVISE] high school graduating class, and by PSAT scores. Two orders indicated PSAT score thresholds ranging from 1070-1180, another two orders filtered for a 1190-1260 PSAT range, and the remaining two orders filtered for scores ranging from 1270-1520.

@fig-la-rq2-race presents the racial/ethnic composition of purchased student profiles. The more than 15,000 prospects in the resulting student lists at the low PSAT score thresholds (1070-1180) reflect very similar racial/ethnic compositions within Geomarkets than overall demographic makeup of the metropolitan area in 2020 (see @fig-la-rq1-race). [THE LEDE IS SOUTH CENTRAL AND EAST LA. START THERE!]. However, the racial/ethnic composition of prospects become less proportional to overall metropolitan patterns by Geomarket as PSAT scores increase. For instance, prospect pools at middle range PSAT scores (1190-1260) within Geomarkets become disproportionately more Asian, while the representation of Hispanic and Black prospects decline substantially. This pattern becomes most pronounced at the highest PSAT score prospect pools. [I DON'T THINK THAT BLACK AND LATINX STUDENTS HAVING LOW TEST SCORES IS THE STORY WE ARE TRYING TO TELL. THAT STORY HAS BEEN TOLD. THE MORE IMPORTANT THING IS, WITHIN SOME TEST THRESHOLD, HOW DOES THE RACIAL COMPOSITION VARY ACROSS GEOMARKETS. LIKE EVEN WITHIN THE MID RANGE, SOUTH CENTRAL, EAST LA, AND INLAND EMPIRE STAND OUT] <!-- This disproportionate over representation for Asian prospects in comparison to the population of residents with Geomarkets ranges from 3 percentage points in the West Los Angeles and West Beach Geomarket (16% prospects versus 13% Asian residents) to 52 percentage points in the East Los Angeles Geomarket (79% prospects versus 27% residents). @fig-la-rq2-race illustrates the share of Black prospects within these high PSAT orders declines to less than 3% within all Geomarkets in comparison to Black residents making up a range of 3% (Covina and West Covina) to 18% (South and South Central Los Angeles) of residents within Geomarkets. Similar underrepresentation patterns are evident for Hispanic porspects are evident across all Geomarkets execept for a nearly equal representation in South and South Central Los Angeles (73% prospects versus 71% Hispanic residents). -->

@fig-la-rq2-midSAT-race presents the race/ethnicity of all prospects targeted in the middle PSAT score range across Geomarket contributions for the Los Angeles metropolitan area. For the more than 2,500 White prospects in this pool, we see the San Fernando Valley - West (28% versus 20% for all prospects) and the West Los Angeles and West Beach (12% versus 7% for all prospects)Geomarkets contributing disproportionately larger shares.

[I THINK CLEANER IF YOU JUST HAVE A SMALL PARAGRAPH FOR HISPANIC AND A SMALL PARAGRAPH FOR BLACK. I THINK ONE LEDE ACROSS THESE RESULTS IS RIVERSIDE AS AN ENGINE FOR BLACK/LATINX AND FOR FIRST-GEN STUDENTS. I GET THAT THESE TWO POINTS CONTRADICT EACH OTHER].
Representational patterns across Geomarkets are reversed for Black and Hispanic prospects in @fig-la-rq2-midSAT-race. South and South Central Los Angeles and Riverside, San Bernadino, and Ontario Geomarkets contributed larger proportions of Black and Hispanic prospects relative to their contributions to the overall prospect pool in the Los Angeles metropolitan area. On the other hand, the San Fernando Valley- West and the Glendale and Pasadena Geomarket contributed smaller proportions of Black and Hispanic Prospects. However, the West Los Angeles and West Beach (10% versus 7% for all prospects), Hollywood Wilshire (7% versus 5% for all prospects), and the Long Beach (8% versus 5% for all prospects) Geomarkets contributed a larger share of Black prospects, whereas Covina and West Covina contributed a larger share of Hispanic students. All other Geomarkets contributed nearly equal shares of Black and Hispanic prospects.

[I MOVED ASIAN] For Asian students, the Glendale and Pasadena (17% versus 14% for all prospects), East Los Angeles (11% versus 5% for all prospects) and Covina and West Covina (17% versus 11% for all prospects) Geomarkets contributed a disproportionate larger shares, whereas nearly all other Geomarkets contributed disproportionately smaller or nearly equal shares.
